---
title: "Lab 2 - Data wrangling"
author: "Derek Dubrule. Student"
date: "2026-02-05"
format: 
  html: default # for speed
  pdf: default # for final output

---


```{r}
#| eval: true
#| message: false

library(tidyverse)
```


# Questions

## Part 1

### Question 1

```{r}
midwest |>
  group_by(state) |>
  summarise(num_counties = n()) |>
  arrange(desc(num_counties))
```

Illinois (IL) has the most amount of counties at 102.

Wisconsin (WI) has the least amount of counties at 72.

\newpage

### Question 2
```{r}
midwest |>
  count(county) |>
  filter(n == n_distinct(midwest$state))
 
```

The counties of CRAWFORD, JACKSON, and MONROE could fill in this blank as these county names appear once in each of the 5 states.

\newpage

### Question 3
```{r}
# Question 3a - population density > 250000
midwest |>
  filter(popdensity > 25000) |>
  select(
    county,
    state,
    popdensity,
    poptotal,
    area
  ) |>
  arrange(desc(popdensity))
```
```{r}
# Question 3b - max population density
midwest |>
  filter(popdensity == max(popdensity)) |>
  select(
    county,
    state,
    popdensity,
    poptotal,
    area
  )
```

\newpage

### Question 5

\newpage

### Question 6

\newpage

### Question 7

\newpage

### Question 8

\newpage

## Part 2


### Question 9

